---
title: "Literate Programming"
author: "Jon Wheeler"
date: "1/19/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome!

Today we will be presenting an introduction to R and RStudio, with a special
focus on writing reproducible reports using R Markdown.

This seminar is intended to serve as a high level introduction to these
applications and concepts. We are planning for longer, more in-depth and
hands-on workshops in the near future.

This notebook is publicly available from GitHub, at 
<https://github.com/unmrds/r-markdown-seminar>. The data are described in more
detail below, but if you would like to run the markdown and R code against
the data used for this demonstration, the data are available using a link
that will be provided in the chat.

You are welcome to launch R or RStudio, download the data, and follow along,
but please note that we may not be able to troubleshoot technical issues 
during our session today. Please do feel free to email us afterwards with 
any questions.

### Today's instructors

* Jon Wheeler, Data Curation Librarian
* Harry Snow, Programmer/Analyst CTSC

### What we will cover

* Overview of R and RStudio
* Introduction to Markdown
* Using community developed statistical packages in R
* Writing parameterized reports

## R and RStudio

Content for this section is adapted from the University Libraries' Research
Data Services Coffee and Code workshop on R and RStudio:
<https://github.com/unmrds/cc-R-RStudio>.

## Markdown

Content for this section is adapted from the University Libraries' Research
Data Services Coffee and Code workshop on Reproducibility and Communication
Using Notebooks (Jupyter (python) and R):
<https://github.com/unmrds/cc-reproducibility-communication>.

## About the Data

Our demonstration today uses data about health related behaviors of
high-school students from the Youth Risk Behavior Surveillance System (YRBSS).
The data are available in multiple formats from the Centers for Disease Control
and Prevention: <https://www.cdc.gov/healthyyouth/data/yrbs/index.htm>.

The dataset we're using is the _Combined YRBS High School Dataset_, which
contains data from every participating high school during each survey year. For
our purposes, the data have been pre-processed a little from the dataset as
provided in MS Access format for states beginning with letters N - Z. 
Specifically, we:

1. Queried the database for NM data for the years 2011, 2013, 2015, 2017, 2019.
2. Exported each year's data in Excel format.

The data are not included in the GitHub repository for this seminar, mainly
because of uncertainty regarding access and use limitations on the data.

### Important!

The remainder of this notebook should not be taken as an actual
analysis of the YRBSS data, but only as a demonstration of the capabilities
of R, RStudio, and R Markdown for building efficiency into analytic and 
reporting workflows.

## Working with the Data in RStudio

Let's start working with the data to produce an example R Markdown report.

The process below is divided into two stages, both of which were included in
our model of a typical R workflow. First, we often need to manipulate, clean,
or otherwise "tidy" data prior to analyzing it. Second, we analyze the data 
and present our findings.

One way to approach this would be to create a separate script for the data
manipulation, and the only include the analysis and visualization steps in our
report. However, R Markdown allows us to specify which code blocks and output
are included in the compiled version of the notebook. This gives us the option
to develop our workflow within a single document.

For example, by setting the "include" argument below to "FALSE," we are able to
exclude the following code block from the compiled report:

```{r load_packages, echo=TRUE, include=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(crosstable)
```

We will take this approach for the rest of this example, leaving out code blocks
that are not part of the analysis and visualization.

Next we read a data file. Note that this step can be parameterized to operate
on a list of files.

```{r read_data, echo=TRUE, include=FALSE, message=FALSE}
fp <- "./data/sadcqn_nm2019.xlsx"
yrbssNM <- read_excel(fp)
```

```{r check_problems, include = FALSE}
# Check the data import for any problems:
problems(yrbssNM)
```

### Tip: adding inline R code

Since we excluded the above code block from the compiled report, we can 
use inline R code to document the name of the file we just read: **`r fp`**.

Note that we are able to use Markdown syntax to print the file name in bold
type. We can also use inline code to print the value of specific cells from a
table:

The value of the first cell of the _weight_ column is:
`r yrbssNM$weight[1]`.

Finally, we can print the output of an operation:

The maximum value of the weight variable is: `r max(yrbssNM$weight)``

When creating reports, this gives us the flexibility to update data points in
our text dynamically as opposed to only being able to update tables and 
figures.


## Incorporating Data and Plots

The data include many categorical variables that have been imported as numeric
and character data types. If we want to perform certain operations or run 
models, we will have to change the data type of these variables to factors. But 
for now we can still interrogate the data as-is and draw a few plots.

Below is a table showing the proportion of survey respondents by sex and grade.
Note that the variables are not labeled - we will add labels below.

```{r proportion_table}
# note that sum of all values is 1
prop.table(table(yrbssNM$grade, yrbssNM$sex))
```

Below is a plot of respondent weights (in kilograms) by grade. Again, the
variables aren't labeled in a way that makes this easy to read but we can
demonstrate an immediate capability of R over Excel. In addition to plotting
variables along the X and Y axes, we are able to include more information in 
the plot by defining aesthetics based on other variables.

```{r plot weight/grade}
ggplot(yrbssNM, mapping = aes(x = grade, y = weight)) + 
         geom_point(alpha = 0.05, aes(color = stheight, size = bmi))
```

## Manipulate the Data

From here on we will work through a process of manipulating the data so that we
can output a report with summary statistics similar to the YRBSS web page.

First we will change some variable data types into factors. Note that we
are excluding all of the following code blocks from our compiled report.

```{r sex_factor, echo = FALSE, message = FALSE, include = FALSE}
# Documentation says A for female, B for male, but we have values of 1 and 2...
# Assume 1 == female, 2 == male

# Note if we want to see the plot below, we need to set include == TRUE
yrbssNM$sex <- as.factor(yrbssNM$sex)
levels(yrbssNM$sex)
levels(yrbssNM$sex)[1] <- "female"
levels(yrbssNM$sex)[2] <- "male"
levels(yrbssNM$sex)
plot(yrbssNM$sex) # omits NAs
```

```{r race4_factor, include = FALSE}
# Same process as above, using a different
# factoring function to make it a one-liner.
yrbssNM$race4 <- factor(yrbssNM$race4,
                            labels = c("White", 
                                       "Black or African American", 
                                       "Hispanic/Latino", 
                                       "All other races"))
levels(yrbssNM$race4)
plot(yrbssNM$race4)
```

```{r race7_factor, include = FALSE}

yrbssNM$race7 <- factor(yrbssNM$race7,
                            labels = c("American Indian/Alaska Native", 
                                       "Asian", 
                                       "Black or African American", 
                                       "Hispanic/Latino",
                                       "Native Hawaiian/Other Pacific Islander",
                                       "White",
                                       "Multiple Races (Non-Hispanic)"))
levels(yrbssNM$race7)
plot(yrbssNM$race7)
```

```{r obesity_factor, include = FALSE}

yrbssNM$qnobese <- factor(yrbssNM$qnobese, labels = c("obese", 
                                                              "not obese"))
levels(yrbssNM$qnobese)
plot(yrbssNM$qnobese)
```


```{r overweight_factor, include=FALSE}
# Factors below for overweight should be treated with special caution, even for
# the purposes of our example. Based on a quick inspection of the the data,
# we don't know that the labels are correct.
yrbssNM$qnowt <- factor(yrbssNM$qnowt, labels = c("overweight", 
                                                          "not overweight"))
levels(yrbssNM$qnowt)
plot(yrbssNM$qnowt)
```

The labels for the factors levels will now be used in tables and plots. If we
redo our proportion table above, we see that values for sex have been
categorized as "female" and "male."

```{r prop_table_mf_factors}
# note that sum of all values is 1
prop.table(table(yrbssNM$grade, yrbssNM$sex))
```
There are other ways to label parts of a plot without having to change the
underlying data types to factors, but note that the labels for our factors
are also used by default in plots.

The plot below could benefit from some reformatting or resizing, but here
again we see the capabilities of R over Excel when it comes to plotting.

```{r plot_race7_bmi_sex}
ggplot(yrbssNM, mapping = aes(x = sex, y = bmi)) + 
         geom_boxplot(aes(fill = qnobese)) + 
         facet_wrap(~ race7) +
         labs(x = "Sex", y = "Body Mass Index", 
              title = "BMI by Race and Sex")
```

```{r sexual_minority_factors, include = FALSE}
# Change sexual minoroty variables to factors.
yrbssNM$q66 <- factor(yrbssNM$q66, labels = c("heterosexual",
                                                      "gay or lesbian",
                                                      "bisexual",
                                                      "not sure"))
yrbssNM$q65 <- factor(yrbssNM$q65, labels = c("never had sexual contact",
                                                      "females",
                                                      "males",
                                                      "females and males"))
yrbssNM$sexid <- factor(yrbssNM$sexid, labels = c("heterosexual",
                                                      "gay or lesbian",
                                                      "bisexual",
                                                      "not sure"))
yrbssNM$sexid2 <- factor(yrbssNM$sexid2, labels = c("heterosexual",
                                                      "sexual minority",
                                                      "unsure"))
yrbssNM$sexpart <- factor(yrbssNM$sexpart, labels = c("never had sex",
                                                      "opposite sex only",
                                                      "same sex only",
                                                      "both sexes"))
yrbssNM$sexpart2 <- factor(yrbssNM$sexpart2, labels = c("never had sex",
                                                      "opposite sex only",
                                                      "same sex only or both sexes"))
```


```{r sexual_minority_plots, include = FALSE}
plot(yrbssNM$q66)
plot(yrbssNM$q65)
plot(yrbssNM$sexid)
plot(yrbssNM$sexid2)
plot(yrbssNM$sexpart)
plot(yrbssNM$sexpart2)
```

## Reporting Statistics on Risk Behavior

The dataset used for this demonstration includes the dichotomized versions
of the remaining questions. 

These variables are set to 1 when the answer is in the response of interest,
and 2 otherwise. More detail is provided in the dataset documentation. 

Variables labeled QN8 - QN11 address driving behaviors. 
For the remainder of this demo we will focus on those variables to derive
the percentage of students who engage in specific behaviors such as not 
wearing a seat belt (QN8).

**QN8:** - _How often do you wear a seat belt when riding in a car
driven by someone else?_ 

Behaviors of interest, per documentation, are responses of "Never" or "Rarely" 
to this question. We can arrive at the percentage of students who engaged in
behaviors of interest:

```{r seat_belt}
# Using dplyr
seat_belts <- yrbssNM %>% 
  select(qn8) %>% 
  filter(!is.na(qn8)) %>% 
  summarise(pct = sum(qn8 == 1)/n())

seat_belts
```

We can break responses down by different demographics. For example, we can
get the distribution of respondent behaviors by sex:

```{r}
seat_belts_mf <- yrbssNM %>% 
  select(sex, qn8) %>% 
  filter(!is.na(qn8), !is.na(sex)) %>% 
  group_by(sex) %>% 
  summarise(c = sum(!is.na(qn8)), 
            c1 = sum(qn8 == 1),
            c2 = sum(qn8 == 2),
            pct = sum(qn8 == 1)/n())

# Note the different ways the below are output between running the code
# versus compiling the document:
seat_belts_mf
ggplot(seat_belts_mf, aes(x = sex, y = pct)) + geom_col()
```
### Tip: Reporting statistics inline 

Returning to our examples of using inline code in R Markdown, depending on our
objectives or audience for a report we can exclude the code and calculations 
from the compiled report and add statistics inline. We will reproduce our
example above to break down seat belt behaviors by age. 

```{r, include = FALSE}
seat_belts_age <- yrbssNM %>% 
  select(age, qn8) %>% 
  filter(!is.na(qn8), !is.na(age)) %>% 
  group_by(age) %>% 
  summarise(c = sum(!is.na(qn8)), 
            c1 = sum(qn8 == 1),
            c2 = sum(qn8 == 2),
            pct = sum(qn8 == 1)/n())
```

Without including our code in the compiled report, we can nonetheless state
that the percentage of respondents 18 or older who never or rarely wear a
seatbelt is `r seat_belts_age$pct[7]`.

We can also still include a plot of the results:

```{r}
ggplot(seat_belts_age, aes(x = age, y = pct)) + geom_col(fill = "blue")
```
## Models

Results of models can also be included in a report. This may be one area where
reporting coefficients using inline code could simplify the presentation.

Keeping in mind that nothing in this example notebook should be taken as an
actual analysis, the following model _may_ predict the likelihood that 
an individual will rarely or never wear a seat belt based on their sex.

Let's exclude the model code block from our compiled report.

```{r, include = FALSE}
yrbssNM$qn8 <- as.factor(yrbssNM$qn8)
yrbssNM$qn11 <- as.factor(yrbssNM$qn11)
sbm <- glm(qn8 ~ sex, data = yrbssNM, family = "binomial")
summary(sbm)
sb_odds <- round(cbind(exp(coef(sbm)),
            exp(confint(sbm))), 3)
sb_odds
```

As noted we can still use the model results in our report. We are going
to avoid interpreting the result, but for the sake of our example
we will suggest that the odds of a randomly selected male always wearing a
seat belt are `r sb_odds[2]`. 

A similar process as we have demonstrated here can be used to produce
descriptive statistics, plots, and models for any of the variables in the
YRBSS data, as well as for cross tabulation of multiple variables, 
multinomial models, etc.

Alth===ough this notebook was developed using the 2019 data, in order to run
this same report against any other year's data we would only have to change
one line of code in the whole report - the value of the `fp` variable above
that we use to point to the raw data file.








