---
title: "YRBSS example report for `r fp`"
author: "Jon Wheeler"
date: "1/25/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Example YRBSS Report for `r fp`

```{r load_packages, echo=TRUE, include=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(crosstable)
```

This report produces the same statistics and plots as we used for our
example report, only the output has been generated per file from a list
of data files.

```{r read_data, echo=TRUE, include=FALSE, message=FALSE}
yrbssNM <- read_excel(paste0("./data/", fp))
```

```{r check_problems, include = FALSE}
# Check the data import for any problems:
problems(yrbssNM)
```

```{r sex_factor, echo = FALSE, message = FALSE, include = FALSE}
# Documentation says A for female, B for male, but we have values of 1 and 2...
# Assume 1 == female, 2 == male

# Note if we want to see the plot below, we need to set include == TRUE
yrbssNM$sex <- as.factor(yrbssNM$sex)
levels(yrbssNM$sex)
levels(yrbssNM$sex)[1] <- "female"
levels(yrbssNM$sex)[2] <- "male"
levels(yrbssNM$sex)
plot(yrbssNM$sex) # omits NAs
```

```{r race4_factor, include = FALSE}
# Same process as above, using a different
# factoring function to make it a one-liner.
yrbssNM$race4 <- factor(yrbssNM$race4,
                            labels = c("White", 
                                       "Black or African American", 
                                       "Hispanic/Latino", 
                                       "All other races"))
levels(yrbssNM$race4)
plot(yrbssNM$race4)
```

```{r race7_factor, include = FALSE}

yrbssNM$race7 <- factor(yrbssNM$race7,
                            labels = c("American Indian/Alaska Native", 
                                       "Asian", 
                                       "Black or African American", 
                                       "Hispanic/Latino",
                                       "Native Hawaiian/Other Pacific Islander",
                                       "White",
                                       "Multiple Races (Non-Hispanic)"))
levels(yrbssNM$race7)
plot(yrbssNM$race7)
```

```{r obesity_factor, include = FALSE}

yrbssNM$qnobese <- factor(yrbssNM$qnobese, labels = c("obese", 
                                                              "not obese"))
levels(yrbssNM$qnobese)
plot(yrbssNM$qnobese)
```


```{r overweight_factor, include=FALSE}
# Factors below for overweight should be treated with special caution, even for
# the purposes of our example. Based on a quick inspection of the the data,
# we don't know that the labels are correct.
yrbssNM$qnowt <- factor(yrbssNM$qnowt, labels = c("overweight", 
                                                          "not overweight"))
levels(yrbssNM$qnowt)
plot(yrbssNM$qnowt)
```

## Descriptive Statistics

Proportion of female and male respondents by grade:

```{r prop_table_mf_factors}
# note that sum of all values is 1
prop.table(table(yrbssNM$grade, yrbssNM$sex))
```

Plot of BMI by sex, with additional information about race and obesity:

```{r plot_race7_bmi_sex}
ggplot(yrbssNM, mapping = aes(x = sex, y = bmi)) + 
         geom_boxplot(aes(fill = qnobese)) + 
         facet_wrap(~ race7) +
         labs(x = "Sex", y = "Body Mass Index", 
              title = "BMI by Race and Sex")
```

## Reporting Statistics on Risk Behavior

The dataset used for this demonstration includes the dichotomized versions
of the remaining questions. 

These variables are set to 1 when the answer is in the response of interest,
and 2 otherwise. More detail is provided in the dataset documentation. 

Variables labeled QN8 - QN11 address driving behaviors. 
For the remainder of this demo we will focus on those variables to derive
the percentage of students who engage in specific behaviors such as not 
wearing a seat belt (QN8).

**QN8:** - _How often do you wear a seat belt when riding in a car
driven by someone else?_ 

Behaviors of interest, per documentation, are responses of "Never" or "Rarely" 
to this question. We can arrive at the percentage of students who engaged in
behaviors of interest:

```{r seat_belt}
# Using dplyr
seat_belts <- yrbssNM %>% 
  select(qn8) %>% 
  filter(!is.na(qn8)) %>% 
  summarise(pct = sum(qn8 == 1)/n())

seat_belts
```

We can break responses down by different demographics. For example, we can
get the distribution of respondent behaviors by sex:

```{r}
seat_belts_mf <- yrbssNM %>% 
  select(sex, qn8) %>% 
  filter(!is.na(qn8), !is.na(sex)) %>% 
  group_by(sex) %>% 
  summarise(c = sum(!is.na(qn8)), 
            c1 = sum(qn8 == 1),
            c2 = sum(qn8 == 2),
            pct = sum(qn8 == 1)/n())

# Note the different ways the below are output between running the code
# versus compiling the document:
seat_belts_mf
ggplot(seat_belts_mf, aes(x = sex, y = pct)) + geom_col()
```

### Tip: Reporting statistics inline 

Returning to our examples of using inline code in R Markdown, depending on our
objectives or audience for a report we can exclude the code and calculations 
from the compiled report and add statistics inline. We will reproduce our
example above to break down seat belt behaviors by age. 

```{r, include = FALSE}
seat_belts_age <- yrbssNM %>% 
  select(age, qn8) %>% 
  filter(!is.na(qn8), !is.na(age)) %>% 
  group_by(age) %>% 
  summarise(c = sum(!is.na(qn8)), 
            c1 = sum(qn8 == 1),
            c2 = sum(qn8 == 2),
            pct = sum(qn8 == 1)/n())
```

Without including our code in the compiled report, we can nonetheless state
that the percentage of respondents 18 or older who never or rarely wear a
seat belt is `r seat_belts_age$pct[7]`.

We can also still include a plot of the results:

```{r}
ggplot(seat_belts_age, aes(x = age, y = pct)) + geom_col(fill = "blue")
```

## Models

Keeping in mind that nothing in this example notebook should be taken as an
actual analysis, our model _may_ predict the likelihood that 
an individual will rarely or never wear a seat belt based on their sex.

```{r, include = FALSE}
yrbssNM$qn8 <- as.factor(yrbssNM$qn8)
yrbssNM$qn11 <- as.factor(yrbssNM$qn11)
sbm <- glm(qn8 ~ sex, data = yrbssNM, family = "binomial")
summary(sbm)
sb_odds <- round(cbind(exp(coef(sbm)),
            exp(confint(sbm))), 3)
sb_odds
```

As noted we can still use the model results in our report. We are going
to avoid interpreting the result, but for the sake of our example
we will suggest that the odds of a randomly selected male always wearing a
seat belt are `r sb_odds[2]`. 

